Ricci-Ollivier curvature of two random walks on rooted phylogenetic subtree-prune-regraft graph

### 2015-02-12 curvature.tex
A bit more intro material.

\usepackage{amsmath}
\usepackage{ifthen}
\usepackage{todonotes}
\usepackage[notref,notcite]{showkeys} % show keys for eqs, etc.

@@

\begin{document}
\title{Ricci-Ollivier curvature of the data-free random walks on rooted phylogenetic subtree-prune-regraft graph}
\author[Whidden]{Christopher Whidden}
\author[Matsen]{Frederick Matsen}
\address{}
\thanks{}

@@

Phylogenetic algorithms do MCMC or heuristic local search.
In both settings, one needs a means of describing "nearby" trees.
There are a number of such
SPR graph.
For MCMC moves, one would like to ensure good mixing.
For ML moves, we are concerned about local optima.
We will call these "graph effects"-- can't get easily from one spot to another because of structure of the graph.
vs. branch length effects?

@@

ML also has graph effects, which manifest as local maxima.
There have been investigation of local maxima, but concern branch lengths.
\cite{Fukami1989-fs,Steel1994-pt,Chor2003-wh,Chor2000-ea}
The work on SPR moves in the ML context has been focused on increasing efficiency of the SPR moves in finding a well-defined maximum (via simulated data) rather than investigating local maxima \cite{Hordijk2005-dl,Stamatakis2005-rh,Guindon2010-lo,Price2010-fi}.

@@

Recently, a new approach to proving theorems about random walks on discrete spaces has been pioneered by Ollivier and colleagues \cite{Ollivier2009-bw,Joulin2010-jg}.

>>>>

\usepackage{amsmath}
\usepackage{ifthen}
\usepackage{todonotes}
\usepackage[notref,notcite]{showkeys} % show keys for eqs, etc.
\usepackage{cite}

@@

\begin{document}
\title[Curvature of phylogenetic subtree-prune-regraft graph]{Ricci-Ollivier curvature of two random walks on rooted phylogenetic subtree-prune-regraft graph}
\author[Whidden]{Christopher Whidden}
\author[Matsen]{Frederick Matsen}
\address{}
\thanks{}

@@

Phylogenetic algorithms do MCMC or heuristic local search.
In both settings, one needs a means of describing "nearby" trees.
There are a number of such means of doing so.
SPR.
SPR graph.
For MCMC moves, one would like to ensure good mixing.
For ML moves, we are concerned about local optima.
We will call these "graph effects"-- can't get easily from one spot to another because of structure of the graph.
We use "random walks" as being a general notion of movement around a graph, including MCMC.

@@

ML also has graph effects, which manifest as local maxima.
There have been investigation of local maxima, but concern branch lengths.
\cite{Fukami1989-fs,Steel1994-pt,Chor2003-wh,Chor2000-ea}
The work on SPR moves in the ML context is primarily found in the literature comparing various SPR variants to one another, usually in the context of comparing one phylogenetic software package to another \cite{Hordijk2005-dl,Stamatakis2006-yz,Price2010-fi,Guindon2010-lo}.
We are not aware of any general conclusions concerning hill-climbing on the SPR or related graphs that has come from this work.

@@

Recently, a new approach to quantifying properties of random walks has been pioneered by Ollivier and colleagues \cite{Ollivier2009-bw,Joulin2010-jg}.
They define a notion of curvature on a discrete or continuous case.
Formalizes idea of the extent to which random walking brings points together, and is generalization of the notion of Ricci curvature on manifolds.

In this paper, we investigate curvature of the rSPR graph with respect to two different random walks and connections to access times for those random walks.
For this first effort we investigate random walks defined in terms of the graph itself: the lazy isotropic random walk and MCMC sampling from the uniform prior on trees.
By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on curvatures under these random walks.
By exploiting symmetries in the rSPR graph, we are able to calculate all of the curvatures for pairs of trees with up to seven leaves.
Using simulation of these processes, we find that the distribution of access times between pairs of trees can be described by distance between the trees, the degrees of the trees, and the curvature.
